\begin{acknowledgements}
This thesis would not have been possible without the help of many people. I am profoundly grateful to my supervisors, Stephen Cameron and Irina Voiculescu, for their guidance, wisdom and (above all) patience. It is only with the benefit of hindsight that I have been able to fully appreciate what was their sometimes rather delphic prescience about what would matter.\footnote{See \emph{Your PhD Companion}, by Stephen Marshall and Nick Green, p.27.}

I am also deeply thankful for the financial support I have received in the course of pursuing my doctorate. In particular, I am indebted to the Engineering and Physical Sciences Research Council (EPSRC) for providing my doctoral training award, and to the Oxford University Computing Laboratory and New College, Oxford, for jointly funding my attendance at international conferences.

My work has been hugely dependent on the goodwill and assistance of the doctors, surgeons, technicians and support staff working in the Churchill Hospital, Oxford. I would especially like to thank:
%
\begin{itemize}

\item Zoe Traill (Consultant Radiologist) for giving up so much of her valuable time to instruct me in the basics of reading CT images, provide me with CT datasets and help validate the results of my algorithms, and Anthony McIntyre (Superintendent Radiographer) for his help in burning CT datasets to disk.

\item David Cranston and Mark Sullivan (Consultant Urological Surgeons) for letting us view operations in order to better understand how they work, and Oliver Dyar (Consultant Anaesthetist) for reviving me when I nearly fainted whilst watching one! (David Cranston also kindly provided some CT datasets.)

\item Andrew Protheroe (Consultant Oncologist) for providing useful feedback and some CT datasets of Von Hippel-Lindau patients.

\item Nilay Patel (Specialist Registrar) for providing CT datasets of partial nephrectomy patients and providing me with an opportunity to test my volume calculations.

\end{itemize}
%
I am also very grateful to Sue Dale, Rachel Shaw, Rosemarie Llewellyn, Carolyn Nelson and Louise Simpson for their help in arranging meetings, and to Professor Sir Mike Brady for his help in setting up our collaboration with the medics.

Last, but by no means least, I would like to thank my girlfriend, Katy Jones, and my mother for their support and encouragement, not to mention making the world a much better place in which to write a thesis.

\end{acknowledgements}
